Explainability and Adversarial Robustness for RNNs

### 2020-02-04 paper/paper.tex
Introduce feature sensitivity to abstract, introduction and conclusion

We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

@@

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently stirred up increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data, but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features, which are manipulated for successful adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.

@@

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow. MLP values from \cite{bachl_walling_2019} are presented for comparison.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r r r} \toprule
& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	& MLP &	Packet	&	Flow & MLP	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.8\%	&	99.5\%	&	98.3\%	&	98.9\%	\\
Precision	&	97.0\%	&	99.7\%	&	99.9\%	&	83.4\%	&	78.6\%	&	84.5\%	\\
Recall	&	97.8\%	&	99.1\%	&	99.2\%	&	87.6\%	&	72.6\%	&	82.9\%	\\
F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
Youden	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
\bottomrule
\end{tabular}
\vspace*{-.3mm}
\end{table}

@@

\section{Defenses}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
\subsection{Reducing Features}
The most obvious defense, which is only possible because of the nature of the dataset, is to simply omit features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving out manipulable features only when the attacker can manipulate them, i.e. only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
\end{itemize}

@@

A question which occurs in this process, is how many CPU cycles to spend on network training and adversarial sample optimization. As a natural approach, we decided to use as many backpropagation steps for neural network training as we use for adversarial sample optimization.  For each adversarial sample optimization step, we performed 10 iterations of gradient descent. Hence, all adversarial samples were optimized each 10 epochs of neural network training.

@@

We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified. We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.

>>>>

We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples. We thus introduce the concept of \textit{feature sensitivity} which quantifies how much potential a feature has to cause misclassification.

Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids} %, capturing a common notion of adversarial robustness, 
and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

@@

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently stirred up increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data, but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features, which are manipulated for successful adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.
Thus, we propose \textit{feature sensitivity} methods, which show how prone a feature is to cause misclassification.

@@

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow. MLP values from \cite{bachl_walling_2019} are presented for comparison.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r r r} \toprule
& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	& MLP &	Packet	&	Flow & MLP	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.8\%	&	99.5\%	&	98.3\%	&	98.9\%	\\
Precision	&	97.0\%	&	99.7\%	&	99.9\%	&	83.4\%	&	78.6\%	&	84.5\%	\\
Recall	&	97.8\%	&	99.1\%	&	99.2\%	&	87.6\%	&	72.6\%	&	82.9\%	\\
F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
Youden	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
\bottomrule
\end{tabular}
%\vspace*{-0.3cm}
\end{table}

@@

\section{Defenses}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
\subsection{Reducing Features}
The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving them out only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which the attacker has control over both sides.
\end{itemize}

@@

A question which occurs in this process, is how many CPU cycles to spend on network training and adversarial sample optimization. We decided to use as many backpropagation steps for training as we use for adversarial sample optimization.  For each adversarial sample optimization step, we performed 10 iterations of gradient descent. Hence, all adversarial samples were optimized each 10 epochs of neural network training.

@@

We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified. Newly developed \textit{Feature sensitivity} methods can show which features can easily be manipulated by an adversary. We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.
